\documentclass[10pt, a4paper]{article}

\usepackage[english]{babel}
\usepackage[margin=2cm, nohead]{geometry}
\usepackage{parskip}
\usepackage{listings}
\usepackage{array}
\usepackage{arydshln}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{shapes,decorations.pathreplacing,positioning,calc,fit}
%\usetikzlibrary{shapes.multipart}


\begin{document}

\begin{center}
\LARGE Block Heap Layout
\end{center}

\section{Abstract}

In this document I will present an alternative memory layout for binary heaps called block heap layout (BHL).
In a $d$-BHL subtrees of height $d$ are stored continuously opposed to the standard layout for binary trees which stores nodes layer by layer.
I will first define the BHL structure and then analyse it in comparison to the standard binary tree layout.

\section{Definition}

The following illustration represents one block in the $d$-BHL structure (the number in the node refers to its position when saved in the array).

\begin{tikzpicture}[level/.style={sibling distance=3cm/#1,level distance=1cm}]
\node [rectangle,draw] (n0){0}
	child {node [rectangle,draw] (n1) {1}
		child {node{$\vdots$}
			child {node [rectangle,draw,left] (n3) {$2^d-1$}}
		}
		child {node{$\vdots$}}
	}
	child {node [rectangle,draw] (n2) {2}
		child {node{$\vdots$}}
		child {node{$\vdots$}
			child {node [rectangle,draw,right] (n4) {$2^{d+1}-2$}}
		}
	};
    
\path (n3) -- (n4) node [midway] {$\cdots$};

\coordinate (cd1) at ($(n4)+(1.5,0)$);
\draw[<->,] 
(cd1) -- (cd1|-n0.east) node [midway,fill=white] {$d$};
\end{tikzpicture}

Let $D=2^{d+1}$. In order to use the BHL structure for heaps of any size, we can make each of the $D$ children of leafs in a block root nodes of another block.
Thus a $D$-ary tree is formed by the blocks.

\begin{tikzpicture}[
	level/.style={sibling distance=3cm/#1,level distance=1cm},
	child anchor=north,parent anchor=south,
	every text node part/.style={align=center},
	nd/.style={isosceles triangle,draw,shape border rotate=90, isosceles triangle apex angle=60,inner sep=0,font=\tiny,yshift={-20mm},text width=2cm},
]
\node[nd] {$0\ldots D-2$}
	child{node[nd,left] {$D\ldots 2D-2$}
		child{node[left] {$\vdots$}}
	}
	child{node[nd,left] (a){$2D\ldots 3D-2$}}
	child{node[nd,right] (b){$D^2\ldots D^2+D-2$}};
\path (a) -- (b) node [midway] {$\cdots$};
\end{tikzpicture}

Note that always one index is skipped between two blocks so that the index of each block's root node is divisible by $D$.

Subsequently, an example for a $1$-BHL layout is depicted:

\begin{tikzpicture}[level/.style={sibling distance=6cm/#1,level distance=1cm}]
\node [rectangle,draw] (n0){0}
	child {node[rectangle, draw] (n1) {1}
		child {node[rectangle, draw] (n4) {4}
			child {node [rectangle, draw] (n5) {5}}
			child {node [rectangle, draw] (n6) {6}}
		}
		child {node[rectangle, draw] (n8) {8}
			child {node [rectangle, draw] (n9) {9}}
			child {node [rectangle, draw] (n10) {10}}
		}
	}
	child {node[rectangle, draw] (n2) {2}
		child {node[rectangle, draw] (n12) {12}
			child {node [rectangle, draw] (n13) {13}}
			child {node [rectangle, draw] (n14) {14}}
		}
		child {node[rectangle, draw] (n16) {16}
			child {node [rectangle, draw] (n17) {17}}
			child {node [rectangle, draw] (n18) {18}}
		}
	};
  
\node[draw,red,dashed,fit=(n0) (n1) (n2),label={[red]0}] {};
\node[draw,red,dashed,fit=(n4) (n5) (n6),label={[red]1}] {};
\node[draw,red,dashed,fit=(n8) (n9) (n10),label={[red]2}] {};
\node[draw,red,dashed,fit=(n12) (n13) (n14),label={[red]3}] {};
\node[draw,red,dashed,fit=(n16) (n17) (n18),label={[red]4}] {};
\end{tikzpicture}

\subsection{Formulas}

In order to obtain the formulas needed to calculate parent and child nodes in the $d$-BHL layout, we can combine the parent and child formulas for binary trees (within a block) and $d$-ary trees (between blocks).

Let $F$ be the number of nonleaf nodes in a block. Then $F=2^D-1$. 

Let $i\ge 0$ be the index of some node. Then $m = i\mod D$ is the index of $i$ within its block and $b = \left\lfloor \frac{i}{D} \right\rfloor$ its block's index. Now we can specify the positions of parent, and children:

\[left(i) = \begin{cases} b\cdot D + 2m + 1, & \textrm{if }m < F \\ D\cdot (b\cdot D + 2(m-F)+1), & \textrm{otherwise} \end{cases}\]

\[right(i) = \begin{cases} b\cdot D + 2m + 2, & \textrm{if }m < F \\ D\cdot (b\cdot D + 2(m-F)+2), & \textrm{otherwise} \end{cases}\]

Let $i>0$. Then $b_p=\left\lfloor\frac{b-1}{D}\right\rfloor$ is the block index of the parent block.

\[parent(i) = \begin{cases} b\cdot D + \left\lfloor\frac{m - 1}{2}\right\rfloor, & \textrm{if }m \neq 0 \\ b_p \cdot D + F - 1 + \left\lfloor\frac{b-b_p\cdot D + 1}{2}\right\rfloor, & \textrm{otherwise} \end{cases}\]

If the tree is filled block by blockwise and each block is filled layerwise then we can obtain the next element's index as follows. Let $n$ be the number of elements in the heap.

\[next(n)=\left\lfloor\frac{n}{D-1}\right\rfloor\cdot D + (n \mod (D-1))\]

\section{Analysis}

With the given formulas a binary heap using BHL can be implemented just like the standard binary heap.

The goal of BHL is to put subtrees of the binary heap into a single block in memory (e.g. a cache line in a CPU or a RAM page).
Since blocks of memory usually have sizes that are a power of 2, we let each block begin at a power of two, i.e. a multiple of $D$.
Thus we have to ensure that each block of the binary heap starts at the beginning of a memory block.

Cache lines in most modern CPUs have a size of 64 bytes. If each key has 4 bytes we can use a 3-BHL layout. We then would have to ensure that the array in which the heap is stored is aligned to a cache line, i.e. its memory address is divisible by 64.

\subsection{Tests}

The following tests were done with multiple compilers for the reason that the optimization described above yielded huge performance increases in GCC 4.8.4 (pre-installed in Linux Mint 17.3) while on newer compilers (clang 3.6.0 was tested as well but the results were very similar to GCC 6.1.1) not so much, which I noticed only quite late.

\begin{Tabular}{r | r | r | r | r | r | r}
	\multicolumn{1}{c}{} & \multicolumn{2}{c}{Normal binary heap} & \multicolumn{2}{c}{Binary heap with 3-BHL} \\
	$n$ & GCC 4.8.4 & GCC 6.1.1 & GCC 4.8.4 & GCC 6.1.1 \\\hline
	$10^6$ & 142439730 & 133711790 & 179537408 & 148117049  \\
	$10^7$ & 3086220360 & 1882807895  & 1866722589 & 1878903331  \\
	$10^8$ & 60736413338 & 26945078895  & 26902074735 & 25765136025  \\
	$10^9$ & 889389255072 & 350881140904  & 342073363689 & 337794208713 \\
\end{Tabular}

We can see that a very small benefit can be gained using BHL with a very large input on the newer compilers. Most interestingly, the performance difference on GCC 4.8.4 is quite high. It seems that newer compiler do some optimization greatly speeding up the normal binary heap implementation.

\subsection{Model}

Let $n$ be the number of elements in the heap, $h$ the height of the binary tree representing the heap. Let $D=2^{d+1}$ be the size of a memory block.

Given a large enough $n$ we can assume for the binary heap that $i$, $parent(i)$ always lie in a different block. Then one \texttt{heapifyDown} operation would need $2h+1$ block accesses.

If we now use $d$-BHL, we would only need $2\left\lfloor\frac{h}{d+1}\right\rfloor-1$ block accesses.
\end{document}